---
title: "Leveraged ETF premium"
output: html_document
---

## Introduction

This is a review of my Master's thesis from 2012 on leveraged ETF premium. 

Here's the original abstract:

> Leveraged exchange-traded funds (LETFs) that allow investors to over-proportionally participate in abstract markets, such as indices, are studied in this paper. LETFs are found to suffer from a performance lag (negative alpha). The lag is categorized in three ways: fund management fees, compounding effects, and a leverage premium. The leverage premium effect intensified during the recent financial crisis (2008/09). In accordance with the approach put forward by Frazzini and Pedersen [2011], various market neutral long/short trading strategies which are sought to exploit the LETF premium are found to have delivered positive abnormal returns over the past years. While these strategies are not necessarily replicable, they teach us a lot about the relation between risk and expected return of LETFs in general.

## Review with R

Hereafter I will review the results using R. The computations in my original thesis were performed using Matlab, but as I'm using R now and there are excellent packages for handling financial data (quantmod, tidyquant) as well as for graphical illustrations (ggplot2), I will be using R.

First, let's load the packaged that we'll use and load the data from Yahoo Finance. We'll use monthly data for now, even though originally in the thesis daily data was used.


```{r setup, message=FALSE}
library(tidyverse)
library(tidyquant)
library(readxl)
library(ggplot2)

etfs <- read_excel("listofetfs.xls")

# prices_d <- tq_get(etfs %>% filter(!terminated) %>% select(Ticker) %>% pull(),
#                    get = "stock.prices",
#                    from = "2000-01-01",
#                    collapse = "monthly")
# 
# write_csv(prices_d, "prices_d.csv")
# 
# prices_m <- prices_d %>% 
#   group_by(symbol) %>%
#   tq_transmute(mutate_fun = to.monthly, indexAt = "lastof")
# 
# write_csv(prices_m, "prices_m.csv")
# 
# rm(prices_d)

prices_m <- read_csv("prices_m.csv")

# prices_d <- read_csv("prices_d.csv")

```

When loading the data from Yahoo Finance we get the error that two symbols are unnavailable: XIV and BGU. XIV was the short-volatility ETF from Credit Suisse that [was terminated in February 2018](https://www.zerohedge.com/news/2018-02-06/credit-suisse-terminates-xiv) after suffering tremendous losses due to spikes in market volatility. The ticker BGU was a tripple-long LETF from Direxion which has also been terminated since 2012.

## Stylized facts

Let's see how the leveraged ETFs have changed in terms of (total) expense ratio (ER).

```{r expense}
etfs %>% filter(!terminated) %>% 
  ggplot() + 
  geom_point(aes(y = ER_2012, x = ER_2018, color = as.factor(abs(Leverage))), size = 3, alpha = 0.8) + 
  geom_abline(aes(intercept = 0, slope = 1), alpha = 0.2) +
  theme_minimal() + labs(y = "ER 2012", x = "ER 2018", color = "Leverage", 
                         title = "(Total) Expense Ratio in 2018 vs. 2012",
                         subtitle = "Dots above the diagonal line represent LETFs that have become more expensive since 2012.",
                         caption = "Data source: etfdb.com")
```

```{r availability}
etfs %>% 
  ggplot() +
  geom_text(aes(x = as.factor(Leverage), y = Index, 
                label = Ticker, color = Issuer, 
                fontface = c("plain", "bold.italic")[terminated+1]), size = 3) +
  theme_minimal() + labs(x = "Leverage factor", y = "Underlying Index", color = "Issuer", 
                         title = "Availability of (L)ETFs per Underlying Index",
                         subtitle = "Ticker symbol, in bold-italic the discontinued LETFs",
                         caption = "Data source: etfdb.com")

```

It is quite remarkable, how ProShares and to some extreme extend Direxion are almost the only issuers of LETFs. Credit Suisse offered it's ETFs on the S&P500 volatility index (VIX), most notable the short-volatility XIV which suffered tremendous losses as volatility spiked in February 2018 and was [terminated shortly after](https://www.zerohedge.com/news/2018-02-06/credit-suisse-terminates-xiv).

Due to unavailability of both *BGZ* and *BGU*, we will exclude the *Russell 1000* entierly from our analysis (only *IWB* remaining, remove!):

```{r removeiwb}
prices_m <- prices_m %>% filter(symbol != "IWB")
```

Let's look at data availability!

```{r timeback}
prices_m %>% group_by(symbol) %>% summarise(first = min(date), last = max(date), obs = n()) %>% 
  left_join(etfs, by = c("symbol" = "Ticker")) %>% select(symbol, first, last, obs, Leverage) %>% 
  ggplot() + 
  geom_linerange(aes(x = reorder(symbol, obs), ymin = first, ymax = last, color = factor(abs(Leverage)))) + 
  coord_flip() +
  theme_minimal() + 
  labs(y = "Availability", x = "Ticker symbol", color = "Leverage", 
                         title = "Availability of (L)ETFs since",
                         subtitle = "",
                         caption = "Data source: etfdb.com, finance.yahoo.com") +
  theme(axis.text=element_text(size=4))
```

[ETFdb.com](www.etfdb.com) lists 70 new double or tripple LETFs incepted since 2013. These include mainly sector specific or regional LETFs, e.g. a 3x Biotech or 3x Mexico, both from Direxion. By the way, it seems that even 4x leveraged ETFs on currency pairs have been made available.

Now let's investigate the popularity. For this we will look at the traded volume.

```{r volume}
prices_d <- read_csv("prices_d.csv")
prices_d <- prices_d %>% filter(symbol != "IWB")

for (set in 1:3) {
print(
prices_d %>% mutate(year = year(date)) %>% group_by(symbol, year) %>% 
  summarise(avgvolume = mean(volume)) %>% 
  left_join(etfs, by = c("symbol" = "Ticker")) %>% select(symbol, year, avgvolume, Ind, Index, Leverage, Set) %>% 
  filter(Set == set) %>% 
  ggplot() + geom_col(aes(y = avgvolume, x = year, fill = factor(Leverage, levels = c(1,-1,2,-2,3,-3))), alpha = 0.7) + 
  facet_wrap(~ Index, scales = "free") + 
  theme_minimal() +
  theme(legend.position="none") +
  labs(y = "avg daily volume", x = "", 
                         title = "Average daily trading volume indicating the popularity",
                         subtitle = str_c("Set ", set),
                         caption = "Data source: finance.yahoo.com")
)
}

remove(prices_d)
```

Besides the spike in traded volume during market turmoils (which we observe in all financial markets), we see that for some indices, the LETFs are more and for some less popular. The simple emerging market ETF *EEM* is highly popular (second highest volume after the *SPY*), while the LETFs on emerging markets cannot keep up with its popularity (relatively). On the other hand we see indices such as the (small cap) Russell 2000, with popular LETFs. Some ETF markets almost vanished alltogether, such as the products on the Dow Jones Financials.

## Data preparation 

First we need some returns. While in academics mostly log-returns are used (also in my original paper), we shall use the more practical arithmetic returns.

```{r returns}
prices_m <- prices_m %>% 
  group_by(symbol) %>% 
  tq_mutate(select = adjusted,
            mutate_fun = periodReturn,
            period = "monthly",
            col_rename = "Ra",
            type="arithmetic")
```

## Trading strategies

In my thesis I tested the original betting-against-beta (*BAB*) strategy from Frazzini and Pedersen (2010). In the most basic BAB strategy, one would short sell a 2x leveraged ETF (e.g. the 2x S&P500 *SSO*) while hedging the trade by purchasing two simple long ETFs on the same underlying index (e.g. the *SPY* in case of the S&P500). I have previously found, that such strategy is profitable, i.e. gives abnormal returns.

```{r simplestrategy}
wgts <- etfs %>% mutate(weight = case_when(Ticker == "SPY" ~ 2,
                                           Ticker == "SSO" ~ -1,
                                           TRUE ~ 0)) %>% 
  filter(!terminated) %>% filter(Ticker != "IWB") %>% 
  select(Ticker, weight)

wgts <- c(
    2, -1,
    1, 0,
    0, 1
)
stocks <- c("SPY", "SSO")
weights_table <-  tibble(stocks) %>%
    tq_repeat_df(n = 3) %>%
    bind_cols(tibble(wgts)) %>%
    group_by(portfolio)
weights_table

wgts %>% mutate(portfolio = 1)

from <- prices_m %>% 
  filter(symbol %in% (wgts %>% filter(weight != 0) %>% select(Ticker) %>% pull())) %>% 
  group_by(symbol) %>% 
  summarise(first = min(date)) %>% 
  ungroup() %>% 
  summarise(first = max(first)) %>% 
  pull()

prices_m %>% filter(date >= from, symbol %in% c("SPY","SSO")) %>%
  tq_repeat_df(n = 3) %>% 
  tq_portfolio(assets_col = symbol,
               returns_col = Ra,
               weights = weights_table,
               col_rename = "return",
               geometric = FALSE,
               wealth.index = TRUE,
               rebalance_on = "months") %>% 
  ggplot() +
  geom_line(aes(x = date, y = return, color = factor(portfolio)))
```

Now let's do this more systematic. In my original work, I laid out 12 different strategies:

```{r strategies}
strategies <- read_excel("strategies.xls") %>% mutate_if(is.numeric , replace_na, replace = 0)

strategies
```

For simplicity, we concentrate on the following 5 strategies for now:

```{r strategies_used}
strategies <- strategies %>% filter(Str %in% c(1,2,4,5,6))
```

Continue with the systematic approach...

```{r systematic}
for (index in etfs %>% select(Index) %>% unique()) {
  print(index)
}
```